Higher-Order unification for free!

### 2024-04-12 paper/main.tex
rework last sec

\section{Other encodings and related work}\label{sec:encodings}

\noindent
One could ignore the similarity between \Uo{} and \Ue{} and ``just''
describe the object language unification procedure in the meta language by
crafting a \elpiIn{unif} routine and using it as follows in
rule \ref{clause:r3}:

\begin{elpicode}
decision X :- unif X (all A x\ app [P, x]), finite A,
  pi x\ decision (app [P, x]).
\end{elpicode}

\noindent
This choice would underuse the logic programming engine provided by
the meta language since, by removing any datum from the head of rules,
indexing degenerates. Moreover the unification procedure \elpiIn{unif}
programmed in the meta language is likely to be an order of magnitude
slower than one that is built-in.

@@

\begin{elpicode}
finite (fin N).
decision (nfact N NF).
decision (all A x\ P x) :- finite A, pi x\ decision (P x).
\end{elpicode}

\noindent
There are two reasons for dismissing this encoding. The first one is that
in CIC it is not always possible to adopt it since the type system
of the meta language is too weak to accommodate for the one of the 
object language. In CIC the lambda abstraction has to carry a type in order to
make type checking decidable. Moreover CIC allows for functions with a
variable arity, like the following example:

\begin{coqcode}
Fixpoint arr T n := if n is S m then T -> arr T m else T.
Definition sum n : arr nat n := ...
Check sum 2   7 8   : nat.
Check sum 3   7 8 9 : nat.
\end{coqcode}

\noindent
The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
The second reason is that the CIC encoding provided by Elpi is used for meta
programming (extending) the Coq system, hence it must accommodate the
manipulation of terms that are now know in advance (not even defined in Coq)
without using introspection primitives such as Prologs's \texttt{functor}
and \texttt{arg}. In this sense constants have to live in an open world, like
the \elpiIn{string} data type used in the examples so far.

@@

\section{Related work and conclusion}

In this paper we show how to lift the meta language higher-order unification
procedure to the object language. Our proposed approach is highly adaptable to
align closely with the behavior of the object language. It is not tightly
coupled with the Coq system but can serve as a flexible framework for meta
programming in any ML.

Our encoding leverages the advantage of not needing to recode the unification
algorithm of the object language. Instead, it utilizes the unification of the
meta language facilitated by the various links we establish to handle
``problematic'' subterms. Additionally, our encoding benefits from the
application of indexing algorithms for static clause filtering.

>>>>

\noindent
\section{Related work and conclusion}

Different strategies can be used to unify terms of the object language in the
meta langauge. One initial approach involves integrating \Uo as a procedure
within the ML, outlined as follows:

\begin{elpicode}
  decision X :- X ~\Uo~(all A x\ app [P, x]), finite A,
    pi x\ decision (app [P, x]).
\end{elpicode}

Opting for this method would result in a suboptimal utilization of the logic
programming engine provided by the meta language, as it degrades indexing by
eliminating data from rule heads. Additionally, embedding \Uo within the meta
language is likely to be significantly slower compared to a built-in
solution.

@@

\begin{elpicode}
decision (all A x\ P x) :- finite A, pi x\ decision (P x).
\end{elpicode}

\noindent
However, two reasons justify rejecting this encoding. Firstly, in CIC, it is
not always feasible to adopt it due to the meta language's type system being
too limited to accommodate that of the object language. \\
Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples.

@@

Our encoding provides a third option that addresses all the considerations
mentioned earlier. It capitalizes on the benefit of not requiring the recoding
of the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
we establish to manage ``problematic'' subterms. Moreover, our encoding takes
advantage of indexing algorithms for static clause filtering. It's noteworthy
that we only replace the minimum necessary information with variables,
specifically targeting \maybeeta and \notllambda subterms.

Our proposed approach is highly adaptable to
align closely with the behavior of the object language. It is not tightly
coupled with the Coq system but can serve as a flexible framework for meta
programming in any ML.
